### 2024-01-22 outputs/paper/datasheet_template.qmd
new files added

---
title: "Analyzing Geographic and Demographic Patterns of Homelessness"
subtitle: "Exploring Spatial Variations in Shelter Demand and Gender/Age-Based Disparities"
author: 
  - Jiwon Choi
thanks: "Code and data are available at: LINK."
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable analysis of Australian politicians. We were unable to find a publicly available dataset in a structured format that had the biographical and political information on Australian politicians that was needed for modelling.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - TBD
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - TBD
4. *Any other comments?*
    - TBD

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- TBD
2. *How many instances are there in total (of each type, if appropriate)?*
	- TBD
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- TBD
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- TBD
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- TBD
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- TBD
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- TBD
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- TBD
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- TBD
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- TBD
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- TBD
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- TBD
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- TBD
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- TBD
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- TBD
16. *Any other comments?*
	- TBD

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- TBD
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- TBD
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- TBD
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- TBD
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- TBD
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- TBD
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- TBD
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- TBD
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- TBD
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
12. *Any other comments?*
	- TBD

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- TBD
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- TBD
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- TBD
4. *Any other comments?*
	- TBD

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- TBD
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- TBD
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- TBD
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- TBD
6. *Any other comments?*
	- TBD

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- TBD
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- TBD
3. *Is there an erratum? If so, please provide a link or other access point.*
	- TBD
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- TBD
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- TBD
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- TBD
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
8. *Any other comments?*
	- TBD

\newpage

# References

### 2024-01-22 outputs/paper/paper.qmd
new files added

---
title: "Analyzing Geographic and Demographic Patterns of Homelessness"
subtitle: "Exploring Spatial Variations in Shelter Demand and Gender/Age-Based Disparities"
author: 
  - Jiwon Choi
thanks: "Code and data are available at [github.com/putrepohere](https://github.com/jmacattack27/ModellingHomelessPopulations)."
date: today
date-format: long
abstract: "This study examines the challenges faced by the homeless despite the presence of shelters, triggered by a recent incident of a homeless person succumbing to severe cold. It investigates geographic and demographic trends in homelessness, analyzing variations in shelter demand based on location and gender/age disparities. Utilizing data from the Shelter Support and Housing Administration division's Shelter Management Information System database, the research forecasts demand and provides critical insights: (1) geographic analysis identifies areas with the highest demand for shelter services, (2) demographic analysis uncovers gender and age-based disparities. These insights empower policymakers, shelter providers, and advocacy groups to allocate resources efficiently and proactively address homelessness challenges, ultimately preventing future hardships and tragedies."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Data Source

The dataset at the core of this research is drawn from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database. This database serves as a comprehensive repository of daily records of active overnight shelter and allied services. It represents the most recent and extensive source of data available for analysis in the context of homelessness support services.

## Data Collection

The data collection process is designed to provide a comprehensive and up-to-date view of shelter and overnight service programs. Key aspects of data collection include:

### Daily Updates

The dataset encompasses daily updated information regarding shelter and overnight service programs. Data is collected and reported on a daily basis, ensuring that the information reflects the most current state of affairs in the shelter system.

### Inclusion of Overnight Service Types

One notable enhancement in the data collection process is the inclusion of various overnight service types. Previously, the dataset primarily focused on traditional shelter programs. However, the revised approach now encompasses all overnight service types for which occupancy data is tracked in the SMIS. This expansion allows for a more comprehensive analysis of the services available to homeless individuals.

### Capacity Categorization

Programs within the dataset are categorized based on their capacity type. This categorization classifies programs as either bed-based or room-based. Bed-based capacity is typically applicable to programs with communal sleeping areas, while room-based capacity is typically associated with family programs and hotel-based programs where individual sleeping rooms are provided. This classification ensures that capacity reporting accurately reflects the nature of the program, preventing over-reporting in room-based programs.

### Dual Capacity Measures

To better evaluate program occupancy rates, the dataset provides two measures of capacity. The first measure, known as funding capacity, reports the intended number of beds or rooms that a program is designed to provide. The second measure, referred to as actual capacity, reflects the number of beds or rooms in service and available for occupancy at the time of reporting. This dual measurement approach enhances the accuracy of assessing program occupancy rates.

Overall, the data collection process is underpinned by a commitment to maintaining accurate and timely records of shelter and overnight service programs. By leveraging this dataset, we gain valuable insights into the dynamic landscape of homelessness, facilitating a more informed and data-driven approach to addressing the needs of homeless individuals.

## Data Analysis

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("outputs/models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-01-22 scripts/00-simulate_data.R

@@ -1,0 +1,64 @@
+#### Preamble ####
+# Purpose: Simulates... [...UPDATE THIS...]
+# Author: Rohan Alexander [...UPDATE THIS...]
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(opendatatoronto)
+library(dplyr)
+library(ggplot2)
+
+# get package
+package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
+package
+
+# get all resources for this package
+resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
+
+# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+
+# load the first datastore resource as a sample
+data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+data
+
+#### Simulate data ####
+simulated_data <-
+  tibble(
+    # Use 1 through to 2703 to represent each division
+    "Sector" = 1:2703,
+    # Randomly pick an option, with replacement, 2703 times
+    "User Group" = sample(
+      x = c("Men", "Women", "Youth", "Mixed Adult", "Families"),
+      size = 2703,
+      replace = TRUE
+    )
+  )
+
+simulated_data
+
+# Count the occurrences of each User Group
+counted_data <- simulated_data %>%
+  count(`User Group`)
+
+# Define custom colors for each User Group
+user_group_colors <- c("Men" = "wheat3", "Women" = "wheat4", "Youth" = "wheat2", 
+                       "Mixed Adult" = "wheat1", "Families" = "yellow4")
+
+# Create a bar plot with custom colors
+plot <- ggplot(counted_data, aes(x = `User Group`, y = n, fill = `User Group`)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values = user_group_colors) +  # Set custom colors
+  labs(x = "User Group", y = "Count", title = "Distribution of User Groups") +
+  theme_minimal() +
[…]

### 2024-01-23 outputs/paper/paper.qmd
major updates on scripts and modified some files

@@
   - Jiwon Choi
-thanks: "Code and data are available at [github.com/putrepohere](https://github.com/jmacattack27/ModellingHomelessPopulations)."
+thanks: "Code and data are available at: [https://github.com/jwonc4602/Analyzing-Geographic-and-Demographic-Patterns-of-Homelessness](https://github.com/jwonc4602/Analyzing-Geographic-and-Demographic-Patterns-of-Homelessness)."
 date: today
@@
 library(palmerpenguins)
+#| message: false
+#| echo: false
+
+#### Workspace set-up ####
+
+#read in cleaned data #
+cleaned_voter_statistics <- read.csv(here::here("inputs/data/cleaned_voter_statistics.csv"))
+show_col_types = FALSE
+
+#read in cleaned data #
+summarized_voter_statistics <- read.csv(here::here("inputs/data/summarized_voter_statistics.csv"))
+show_col_types = FALSE
+
+#read in cleaned data #
+summarized_voter_turnout <- read.csv(here::here("inputs/data/summarized_voter_turnout.csv"))
+show_col_types = FALSE
+
+#read in cleaned data #
+finalized_voter_statistics <- read.csv(here::here("inputs/data/finalized_voter_statistics.csv"))
+show_col_types = FALSE
 ```

### 2024-01-23 scripts/00-simulate_data.R

@@ -1,64 +1,121 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulate the 2024 DAILY SHELTER & OVERNIGHT SERVICE OCCUPANCY & CAPACITY statistics data set
+# Author: Jiwon Choi
+# Date: 22 January 2024
+# Contact: jwon.choi@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None
+# Data sets:https://open.toronto.ca/dataset/daily-shelter-overnight-service-occupancy-capacity/
 
 
 #### Workspace setup ####
 library(tidyverse)
+library(ggplot2)
 library(opendatatoronto)
 library(dplyr)
-library(ggplot2)
 
-# get package
-package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
-package
+#### Simulation 1) Distribution of User Groups ####
+set.seed(1250)  # Set a specific seed value for reproducibility
 
-# get all resources for this package
-resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
-
-# load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
-data
-
-#### Simulate data ####
-simulated_data <-
+simulated_data1 <-
   tibble(
     # Use 1 through to 2703 to represent each division
-    "Sector" = 1:2703,
+    "Sector" = 1:2839,
     # Randomly pick an option, with replacement, 2703 times
     "User Group" = sample(
       x = c("Men", "Women", "Youth", "Mixed Adult", "Families"),
-      size = 2703,
+      size = 2839,
       replace = TRUE
     )
   )
 
-simulated_data
+simulated_data1
 
 # Count the occurrences of each User Group
-counted_data <- simulated_data %>%
+counted_data1 <- simulated_data1 %>%
   count(`User Group`)
[…]

### 2024-01-23 scripts/01-download_data.R

@@ -1,26 +1,36 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads, saves, and reads data from the OpenDataToronto portal
+# Author: Jiwon Choi
+# Date: 22 January 2024
+# Contact: jwon.choi@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: None
+# Data sets: https://open.toronto.ca/dataset/daily-shelter-overnight-service-occupancy-capacity/
 
 #### Workspace setup ####
 library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
+# Replace with the specific package ID for your dataset
+package_id <- "21c83b32-d5a8-4106-a54f-010dbe49f6f2"
 
+# Get the package
+package <- show_package(package_id)
+package
 
+# Get all resources for this package
+resources <- list_package_resources(package_id)
 
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+# Identify datastore resources; Toronto Open Data sets datastore resource format to CSV for non-geospatial
+datastore_resources <- filter(resources, tolower(format) %in% c('csv'))
 
-         
+# Load the first datastore resource as a sample
+data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
+
+# Save data
+write_csv(raw_shelter_data, "inputs/data/raw_shelter_statistics.csv")  # Adjust the file path and name
+
+# Read in the DAILY SHELTER & OVERNIGHT SERVICE OCCUPANCY & CAPACITY statistics data set
+raw_shelter_data <- read_csv("inputs/data/raw_shelter_statistics.csv")
+

### 2024-01-23 scripts/02-data_cleaning.R

@@ -1,44 +1,50 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans and prepares the Daily Shelter & Overnight Service Occupancy & Capacity statistics data 
+#          for analysis. This script focuses on data downloaded from the OpenDataToronto portal, ensuring 
+#          it's in the correct format and ready for further statistical analysis or visualization.
+# Author: Jiwon Choi
+# Date: 22 January 2024
+# Contact: jwon.choi@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 01-download.R (Script for downloading the data set from OpenDataToronto)
 
 #### Workspace setup ####
 library(tidyverse)
+library(tidyr)
+library(janitor)
 
-#### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+#### Read and Process Data ####
+# Read in the downloaded shelter data
+raw_shelter_data <- read_csv("inputs/data/raw_shelter_statistics.csv")
 
-cleaned_data <-
-  raw_data |>
+# Process the data
+cleaned_shelter_data <- raw_shelter_data |>
   janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
+  select(occupancy_date, sector, capacity_type, location_city, occupied_beds, capacity_actual_bed, occupied_rooms, capacity_actual_room) |>
   mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
[…]

### 2024-01-23 outputs/paper/paper.qmd
the first submission, render qmd to pdf

@@
 #| message: false
-
-library(tidyverse)
-library(palmerpenguins)
-#| message: false
-#| echo: false
-
-#### Workspace set-up ####
-
-#read in cleaned data #
-cleaned_voter_statistics <- read.csv(here::here("inputs/data/cleaned_voter_statistics.csv"))
-show_col_types = FALSE
-
-#read in cleaned data #
-summarized_voter_statistics <- read.csv(here::here("inputs/data/summarized_voter_statistics.csv"))
-show_col_types = FALSE
-
-#read in cleaned data #
-summarized_voter_turnout <- read.csv(here::here("inputs/data/summarized_voter_turnout.csv"))
-show_col_types = FALSE
-
-#read in cleaned data #
-finalized_voter_statistics <- read.csv(here::here("inputs/data/finalized_voter_statistics.csv"))
-show_col_types = FALSE
 ```
@@
 #| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
 ```

### 2024-01-23 outputs/paper/paper.qmd
major updates on data analysis part of paper

@@
 date-format: long
-abstract: "This study examines the challenges faced by the homeless despite the presence of shelters, triggered by a recent incident of a homeless person succumbing to severe cold.
-It investigates geographic and demographic trends in homelessness, analyzing variations in shelter demand based on location and gender/age disparities.
-Utilizing data from the Shelter Support and Housing Administration division's Shelter Management Information System database, the research forecasts demand and provides critical insights: (1) geographic analysis identifies areas with the highest demand for shelter services, (2) demographic analysis uncovers gender and age-based disparities.
-These insights empower policymakers, shelter providers, and advocacy groups to allocate resources efficiently and proactively address homelessness challenges, ultimately preventing future hardships and tragedies."
+abstract: "This research investigates the difficulties encountered by the homeless, even with available shelters, spurred by a recent case of a homeless individual dying due to extreme cold.
+By conducting geographic and demographic assessments of shelter needs, using data from the Shelter Management Information System, it pinpoints areas with significant demand and uncovers disparities in shelter use based on gender and age.
+The results assist policymakers, shelter operators, and advocacy groups in effectively distributing resources and actively addressing homelessness issues."
 format: pdf
+toc: true
 number-sections: true
@@
 #| message: false
+
+#### Workspace set-up ####
+
+## Load packages
+library(tidyverse)
+library(knitr)
+library(here)
+library(dplyr)
+library(ggplot2)
+library(readr)
+
+#read in cleaned data #
+cleaned_shelter_statistics <- read.csv(here::here("outputs/data/cleaned_shelter_statistics.csv"))
+show_col_types = FALSE
+
+#read in cleaned data #
+finalized_shelter_statistics <- read.csv(here::here("outputs/data/finalized_shelter_statistics.csv"))
+show_col_types = FALSE
 ```
@@
 
-The dataset at the core of this research is drawn from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database.
-This database serves as a comprehensive repository of daily records of active overnight shelter and allied services.
-It represents the most recent and extensive source of data available for analysis in the context of homelessness support services.
+This research is based on a dataset sourced from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto].
+The specific dataset employed is named 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData].
+This database acts as a detailed collection of daily entries related to active overnight shelter and associated services.
+It is the latest and most comprehensive dataset available for analyzing homelessness support services.
 
 ## Data Collection
+In the process of selecting an appropriate data source for an in-depth analysis of current demographic and geographic trends within shelter services, I evaluated several datasets, including the 'Toronto Shelter System Flow data' [@ShelterSystemFlow] and the 'Daily Shelter Occupancy' [@DailyShelterOccupancy] datasets.
+However, I ultimately chose the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset for several compelling reasons:
 
-The data collection process is designed to provide a comprehensive and up-to-date view of shelter and overnight service programs. Key aspects of data collection include:
+1.
+While the 'Toronto Shelter System Flow' [@ShelterSystemFlow] data offers valuable insights into the flow of demography through the shelter system, it lacks comprehensive geographic information.
+This limitation significantly hinders the potential for a detailed analysis of geographic trends and patterns within the shelter system, an aspect critical to my research objectives.
 
-### Daily Updates
+2.
+The 'Daily Shelter Occupancy' [@DailyShelterOccupancy] dataset, although previously a pertinent source of information, has become less relevant due to its discontinuation.
+The most recent update for this dataset was recorded on March 10, 2022, rendering it outdated for current analysis purposes.
+In contrast, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset provides the most recent and comprehensive data available, making it a more suitable choice for analyzing up-to-date trends and developments in shelter usage and capacity.
 
-The dataset encompasses daily updated information regarding shelter and overnight service programs.
-Data is collected and reported on a daily basis, ensuring that the information reflects the most current state of affairs in the shelter system.
+Consequently, the decision to utilize the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset was driven by its superior relevance, contemporaneity, and the inclusion of both demographic and geographic data, which are essential for a thorough and current analysis of shelter services.
 
-### Inclusion of Overnight Service Types
-
-One notable enhancement in the data collection process is the inclusion of various overnight service types.
-Previously, the dataset primarily focused on traditional shelter programs.
-However, the revised approach now encompasses all overnight service types for which occupancy data is tracked in the SMIS.
-This expansion allows for a more comprehensive analysis of the services available to homeless individuals.
-
-### Capacity Categorization
-
-Programs within the dataset are categorized based on their capacity type.
-This categorization classifies programs as either bed-based or room-based.
-Bed-based capacity is typically applicable to programs with communal sleeping areas, while room-based capacity is typically associated with family programs and hotel-based programs where individual sleeping rooms are provided.
-This classification ensures that capacity reporting accurately reflects the nature of the program, preventing over-reporting in room-based programs.
-
-### Dual Capacity Measures
-
-To better evaluate program occupancy rates, the dataset provides two measures of capacity.
-The first measure, known as funding capacity, reports the intended number of beds or rooms that a program is designed to provide.
-The second measure, referred to as actual capacity, reflects the number of beds or rooms in service and available for occupancy at the time of reporting.
-This dual measurement approach enhances the accuracy of assessing program occupancy rates.
-
-Overall, the data collection process is underpinned by a commitment to maintaining accurate and timely records of shelter and overnight service programs.
-By leveraging this dataset, we gain valuable insights into the dynamic landscape of homelessness, facilitating a more informed and data-driven approach to addressing the needs of homeless individuals.
-
-## Data Analysis
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-```{=tex}
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-```
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
+The methodology for gathering data is structured to ensure a detailed and current perspective of shelter and overnight service initiatives.
+The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `knitr` [@rknitr], and `here` [@rHere].
 
 ```{r}
+#| message: false
 #| echo: false
-#| eval: true
-#| warning: false
-#| message: false
+#| label: tbl-cleaned_shelter_statistics
+#| tbl-cap: Sample of Cleaned Shelter Statistics Data
 
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("outputs/models/first_model.rds"))
+head (finalized_shelter_statistics, 5) |>
+  kable(
+    col.names = c("occupancy_date", "sector", "capacity_type",	"location_city", "occupancy_rate"),
+  booktabs = TRUE
+  )
 ```
 
+## Trends in Shelter Occupancy Rates by Dates
+```{r fig.width = 15, fig.height = 8, fig.align = "center", fig.pos = 'h'}
+#| message: false
+#| echo: false
+#| label: fig-average-occupancy-rate-by-date
+#| fig-cap: Average Occupancy Rate by Date
+
+# Assuming 'finalized_shelter_statistics.csv' is your file path
+occupancy_data <- finalized_shelter_statistics
+
+# Calculate the average occupancy rate by date
+average_occupancy_rates <- occupancy_data %>%
+  group_by(occupancy_date) %>%
+  summarise(average_occupancy_rate = sum(occupancy_rate, na.rm = TRUE) / n(), .groups = 'drop')
+
+# Print the results
+# print(average_occupancy_rates)
+
+ggplot(average_occupancy_rates, aes(x = occupancy_date, y = average_occupancy_rate, group = 1)) +
+  geom_line(color = "red") +
+  labs(title = "Average Occupancy Rate by Date",
+       x = "Date",
+       y = "Average Occupancy Rate") +
+  theme(
+    plot.title = element_text(size = 20, hjust = 0.5),          # Center and set size for plot title
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 17), # Set size for x-axis text
+    axis.text.y = element_text(hjust = 1, size = 17),            # Set size for y-axis text
+    axis.title.x = element_text(size = 20, hjust = 0.5),         # Center and set size for x-axis title
+    axis.title.y = element_text(size = 20, hjust = 0.5),         # Center and set size for y-axis title
+  )
+
+```
+The latest analysis of data regarding shelter occupancy shows a significant rising trend in average occupancy rates across a span of 21 days, starting from January 1st and ending on January 21st.
+While there were some daily variations in occupancy rates, the general trend observed was a consistent rise in the need for shelter services.
+This overarching pattern is further illustrated in Figure 1.
+
+- Starting Point (January 1st): The average occupancy rate on the first of January was 97.30062%.
+This figure represents a robust utilization of shelter resources, indicating a high demand right at the start of the year.
+
+- Intermediate Fluctuations: Throughout the first three weeks of January, there were instances of fluctuating occupancy rates.
+These fluctuations, however, did not significantly deviate from the overall upward trajectory.
+Some days witnessed a decrease in occupancy rates, which can be attributed to various transient factors such as weather conditions, temporary housing solutions, or individual decisions to seek alternative accommodations.
+
+- End Point (January 21st): By the 21st of January, the average occupancy rate had risen to 98.33322%.
+This incremental increase, while seemingly modest, is statistically significant and points to a growing demand for shelter services.
+
+## Capacity Types in Shelter Services
 ```{r}
+#| message: false
 #| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+#| label: fig-number-of-capacity-type-by-user-group
+#| fig-cap: Number of Capacity Type by User Group
+#| 
+number_of_capacity_type_by_user_group <- finalized_shelter_statistics
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
+# Group the data by sector and capacity_type, and count the number of occurrences
+grouped_data <- number_of_capacity_type_by_user_group %>%
+  group_by(sector, capacity_type) %>%
+  summarise(count = n())
+
+# Convert factors to ensure proper ordering in the plot
+grouped_data$sector <- factor(grouped_data$sector, levels = unique(number_of_capacity_type_by_user_group$sector))
+grouped_data$capacity_type <- factor(grouped_data$capacity_type, levels = c("Bed Based Capacity", "Room Based Capacity"))
+
+# Generate the bar chart
+ggplot(grouped_data, aes(x = sector, y = count, fill = capacity_type)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
+  theme_minimal() +
+  labs(y = "Count", x = "User Group", fill = "Capacity Type") +
+  scale_fill_manual(values = c("Bed Based Capacity" = "green", "Room Based Capacity" = "orange")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for better readability
+
+```
+A focused analysis of the capacity types utilized in shelter services reveals distinct patterns in the allocation and utilization of resources, particularly when comparing bed-based and room-based capacities across different user groups.
+
+- Absence of Bed-Based Capacity for Families: A striking observation is the complete lack of bed-based capacity options for family groups.
+This suggests that the current shelter system does not offer bed-specific accommodations for families, possibly due to the unique needs and privacy concerns associated with family units.
+This gap in service provision indicates a potential area for development and improvement in catering to the needs of families seeking shelter.
+
+- Dominance of Bed-Based Capacity for Other User Groups: In contrast to the situation with families, other user groups predominantly utilize bed-based capacity in shelters.
+This trend is significant, indicating that for individuals and non-family groups, bed-based accommodations are the primary form of shelter provided.
+This form of capacity is likely favored due to its efficiency in maximizing space and accommodating a larger number of individuals within the shelter system.
+
+- Limited Use of Room-Based Capacity: The analysis also shows that room-based capacity, while available, is less commonly used than bed-based options for non-family groups.
+This could be due to various factors, including limited availability, higher maintenance costs, or specific shelter policies and user preferences.
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-average-occupancy-rate-by-city
+#| fig-cap: Average Occupancy Rate by City
+#| 
+data <- finalized_shelter_statistics
+# Calculate the average occupancy rate by city
+average_occupancy_by_city <- data %>%
+  group_by(location_city) %>%
+  summarise(average_occupied_rate_by_city = mean(occupancy_rate))
+
+# Create a bar plot
+ggplot(average_occupancy_by_city, aes(x = location_city, y = average_occupied_rate_by_city)) +
+  geom_bar(stat = 'identity') +
+  theme_minimal() +
+  labs(title = 'Average Occupancy Rate by City',
+       x = 'City',
+       y = 'Average Occupancy Rate') +
+  ylim(0, max(average_occupancy_by_city$average_occupied_rate_by_city) + 10)
+
+# Note: Users should replace 'your_data.csv' with the actual path to their data file.
+
 ```
@@
 
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix {.unnumbered}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
 \newpage

### 2024-01-25 outputs/paper/paper.qmd
Final revise

@@
 title: "Analyzing Geographic and Demographic Patterns of Homelessness"
-subtitle: "Exploring Spatial Variations in Shelter Demand and Gender/Age-Based Disparities"
+subtitle: "Exploring Spatial Variations in Homeless Shelter Demand and Gender/Age-Based Disparities"
 author: 
@@
 date-format: long
-abstract: "This research investigates the difficulties encountered by the homeless, even with available shelters, spurred by a recent case of a homeless individual dying due to extreme cold.
-By conducting geographic and demographic assessments of shelter needs, using data from the Shelter Management Information System, it pinpoints areas with significant demand and uncovers disparities in shelter use based on gender and age.
+abstract: "This research investigates the difficulties encountered by the homeless, even with available shelters, motivated by a recent case [@news1] of a homeless individual dying due to extreme cold.
+By conducting geographic and demographic patterns of shelter needs, using data from the Shelter Management Information System, it pinpoints areas with significant demand and uncovers trends in shelter use based on gender and age.
 The results assist policymakers, shelter operators, and advocacy groups in effectively distributing resources and actively addressing homelessness issues."
@@
 # Introduction
+Homelessness in Toronto presents a challenging and dynamic issue, influenced by demographic shifts, economic factors, and policy decisions.
+Despite a robust shelter system, designed to protect the city's most vulnerable, instances of individuals not benefiting from these services and facing tragic outcomes still occur.
+This paper aims to analyze demographic and geographic trends in shelter systems to leverage them to resolve and support future homeless shelter issues in Toronto.
 
-You can and should cross-reference sections and sub-sections.
+This study is designed for an in-depth analysis of Toronto's shelter occupancy and capacity, utilizing data [@ShelterData] from the Shelter Management Information System (SMIS) database.
+While there are extensive papers on homelessness, a clear gap exists in contemporary, data-driven analyses that integrate both demographic and geographic dimensions within the context of urban shelter systems.
+Previous studies have often been limited either in data scope or timeframe, failing to provide a comprehensive picture that reflects the current realities of urban homelessness.
+This paper addresses this gap better by offering an updated view of the shelter system in Toronto.
 
-The remainder of this paper is structured as follows. @sec-data....
+This study analyzes shelter occupancy trends using a dataset [@ShelterData] from the Shelter Management Information System (SMIS) [@ShelterData], employing R [@r] statistical tools like `tidyverse` [@rTidyverse] and `ggplot2` [@rGgplot2].
+Findings reveal a heavy reliance on bed-based capacities mainly for non-family groups but not for families, alongside high occupancy rates across Toronto.
+These results are vital for guiding policymakers and social welfare organizations in enhancing shelter capacities and accessibility, and in reevaluating strategies to more effectively address urban homelessness.
+
+To dissect the demographic and geographic patterns within the shelter system in Toronto, spotlighting trends in service utilization and capacity, this paper is organized into the following sections: Data, and Discussion.
+In the Data section, I discuss a detailed description of the data source and collection methods.
+It then progresses to an analysis of demographic and geographic patterns in shelter usage, followed by a discussion on the observed trends and their broader implications.
+The final sections of the discussion part critically evaluate the study's limitations and propose directions for future research.
 
@@
 This research is based on a dataset sourced from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto].
-The specific dataset employed is named 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData].
+The specific dataset employed is named  [@ShelterData].
 This database acts as a detailed collection of daily entries related to active overnight shelter and associated services.
@@
 ## Data Collection
-In the process of selecting an appropriate data source for an in-depth analysis of current demographic and geographic trends within shelter services, I evaluated several datasets, including the 'Toronto Shelter System Flow data' [@ShelterSystemFlow] and the 'Daily Shelter Occupancy' [@DailyShelterOccupancy] datasets.
-However, I ultimately chose the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset for several compelling reasons:
+In the process of selecting an appropriate data source for an in-depth analysis of current demographic and geographic trends within shelter services, several datasets were considered, including the 'Toronto Shelter System Flow data' [@ShelterSystemFlow] and the 'Daily Shelter Occupancy' [@DailyShelterOccupancy] datasets.
+However, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset was ultimately chosen for the following reasons:
 
@@
 The 'Daily Shelter Occupancy' [@DailyShelterOccupancy] dataset, although previously a pertinent source of information, has become less relevant due to its discontinuation.
-The most recent update for this dataset was recorded on March 10, 2022, rendering it outdated for current analysis purposes.
+The most recent update for this dataset was recorded on March 10, 2022, therefore it is outdated for current analysis purposes.
 In contrast, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset provides the most recent and comprehensive data available, making it a more suitable choice for analyzing up-to-date trends and developments in shelter usage and capacity.
 
-Consequently, the decision to utilize the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset was driven by its superior relevance, contemporaneity, and the inclusion of both demographic and geographic data, which are essential for a thorough and current analysis of shelter services.
+Consequently, the decision to utilize the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset was driven by its superior relevance, up-to-date, and the inclusion of both demographic and geographic data, which are essential for a thorough and current analysis of shelter services.
 
 The methodology for gathering data is structured to ensure a detailed and current perspective of shelter and overnight service initiatives.
-The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `knitr` [@rknitr], and `here` [@rHere].
+The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `knitr` [@rknitr], and `here` [@rHere] (see @tbl-cleaned-shelter-statistics).
 
@@
 #| echo: false
-#| label: tbl-cleaned_shelter_statistics
+#| label: tbl-cleaned-shelter-statistics
 #| tbl-cap: Sample of Cleaned Shelter Statistics Data
@@
   kable(
-    col.names = c("occupancy_date", "sector", "capacity_type",	"location_city", "occupancy_rate"),
+    col.names = c("occupancy_date", "sector", "capacity_type",	"location_city", "occupancy_rate(%)"),
   booktabs = TRUE
@@
 
-## Trends in Shelter Occupancy Rates by Dates
-```{r fig.width = 15, fig.height = 8, fig.align = "center", fig.pos = 'h'}
-#| message: false
-#| echo: false
-#| label: fig-average-occupancy-rate-by-date
-#| fig-cap: Average Occupancy Rate by Date
-
-# Assuming 'finalized_shelter_statistics.csv' is your file path
-occupancy_data <- finalized_shelter_statistics
-
-# Calculate the average occupancy rate by date
-average_occupancy_rates <- occupancy_data %>%
-  group_by(occupancy_date) %>%
-  summarise(average_occupancy_rate = sum(occupancy_rate, na.rm = TRUE) / n(), .groups = 'drop')
-
-# Print the results
-# print(average_occupancy_rates)
-
-ggplot(average_occupancy_rates, aes(x = occupancy_date, y = average_occupancy_rate, group = 1)) +
-  geom_line(color = "red") +
-  labs(title = "Average Occupancy Rate by Date",
-       x = "Date",
-       y = "Average Occupancy Rate") +
-  theme(
-    plot.title = element_text(size = 20, hjust = 0.5),          # Center and set size for plot title
-    axis.text.x = element_text(angle = 45, hjust = 1, size = 17), # Set size for x-axis text
-    axis.text.y = element_text(hjust = 1, size = 17),            # Set size for y-axis text
-    axis.title.x = element_text(size = 20, hjust = 0.5),         # Center and set size for x-axis title
-    axis.title.y = element_text(size = 20, hjust = 0.5),         # Center and set size for y-axis title
-  )
-
-```
-The latest analysis of data regarding shelter occupancy shows a significant rising trend in average occupancy rates across a span of 21 days, starting from January 1st and ending on January 21st.
-While there were some daily variations in occupancy rates, the general trend observed was a consistent rise in the need for shelter services.
-This overarching pattern is further illustrated in Figure 1.
-
-- Starting Point (January 1st): The average occupancy rate on the first of January was 97.30062%.
-This figure represents a robust utilization of shelter resources, indicating a high demand right at the start of the year.
-
-- Intermediate Fluctuations: Throughout the first three weeks of January, there were instances of fluctuating occupancy rates.
-These fluctuations, however, did not significantly deviate from the overall upward trajectory.
-Some days witnessed a decrease in occupancy rates, which can be attributed to various transient factors such as weather conditions, temporary housing solutions, or individual decisions to seek alternative accommodations.
-
-- End Point (January 21st): By the 21st of January, the average occupancy rate had risen to 98.33322%.
-This incremental increase, while seemingly modest, is statistically significant and points to a growing demand for shelter services.
-
-## Capacity Types in Shelter Services
+## Demographic Patterns of Homelessness
 ```{r}
@@
 #| fig-cap: Number of Capacity Type by User Group
-#| 
+
 number_of_capacity_type_by_user_group <- finalized_shelter_statistics
@@
 ```
+
 A focused analysis of the capacity types utilized in shelter services reveals distinct patterns in the allocation and utilization of resources, particularly when comparing bed-based and room-based capacities across different user groups.
+Bed-based capacity is typically applicable for programs with common sleeping areas, while room-based capacity is measured at the room level where sleeping rooms are not shared by people from different households (see @fig-number-of-capacity-type-by-user-group).
 
-- Absence of Bed-Based Capacity for Families: A striking observation is the complete lack of bed-based capacity options for family groups.
+A striking observation is the complete lack of bed-based capacity options for family group.
 This suggests that the current shelter system does not offer bed-specific accommodations for families, possibly due to the unique needs and privacy concerns associated with family units.
-This gap in service provision indicates a potential area for development and improvement in catering to the needs of families seeking shelter.
 
-- Dominance of Bed-Based Capacity for Other User Groups: In contrast to the situation with families, other user groups predominantly utilize bed-based capacity in shelters.
+In contrast to the situation with families, other user groups predominantly utilize bed-based capacity in shelters.
 This trend is significant, indicating that for individuals and non-family groups, bed-based accommodations are the primary form of shelter provided.
@@
 
-- Limited Use of Room-Based Capacity: The analysis also shows that room-based capacity, while available, is less commonly used than bed-based options for non-family groups.
-This could be due to various factors, including limited availability, higher maintenance costs, or specific shelter policies and user preferences.
+## Geographic Patterns of Homelessness
 ```{r}
@@
 #| fig-cap: Average Occupancy Rate by City
-#| 
-data <- finalized_shelter_statistics
+
+# Assuming 'location_data' holds the dataset
+location_data <- finalized_shelter_statistics
+
 # Calculate the average occupancy rate by city
-average_occupancy_by_city <- data %>%
+average_occupancy_by_city <- location_data %>%
   group_by(location_city) %>%
@@
 
-# Create a bar plot
+# Create a bar plot with numbers on bars
 ggplot(average_occupancy_by_city, aes(x = location_city, y = average_occupied_rate_by_city)) +
   geom_bar(stat = 'identity') +
+  geom_text(aes(label = round(average_occupied_rate_by_city, 1)), 
+            vjust = -0.5, # Adjust this to position the text above the bars
+            size = 3) + # Adjust text size as needed
   theme_minimal() +
@@
        x = 'City',
-       y = 'Average Occupancy Rate') +
+       y = 'Average Occupancy Rate (%)') +
   ylim(0, max(average_occupancy_by_city$average_occupied_rate_by_city) + 10)
+```
 
-# Note: Users should replace 'your_data.csv' with the actual path to their data file.
+Here are occupancy rates, rounded up at the first decimal place, across six categories: five cities—Etobicoke, North York, Scarborough, Toronto, Vaughan—and an unspecified category labeled "NA" (see @fig-average-occupancy-rate-by-city).
+
+The narrow range between the highest and lowest occupancy rates and such high rates could imply limited vacancies and potential competitiveness in occupancy for homeless people seeking accommodation.
+For policymakers, shelter operators, and advocacy groups, these figures might signal the need for increased shelter supply.
+
+# Discussion
+
+## Trends in Shelter Occupancy Rates by Dates
+```{r fig.width = 15, fig.height = 8, fig.align = "center", fig.pos = 'h'}
+#| message: false
+#| echo: false
+#| label: fig-average-occupancy-rate-by-date
+#| fig-cap: Average Occupancy Rate by Date
+
+# Assuming 'finalized_shelter_statistics.csv' is your file path
+occupancy_data <- finalized_shelter_statistics
+
+# Calculate the average occupancy rate by date
+average_occupancy_rates <- occupancy_data %>%
+  group_by(occupancy_date) %>%
+  summarise(average_occupancy_rate = sum(occupancy_rate, na.rm = TRUE) / n(), .groups = 'drop')
+
+ggplot(average_occupancy_rates, aes(x = occupancy_date, y = average_occupancy_rate, group = 1)) +
+  geom_line(color = "red") +
+  labs(title = "Average Occupancy Rate by Date",
+       x = "Date",
+       y = "Average Occupancy Rate (%)") +
+  theme(
+    plot.title = element_text(size = 20, hjust = 0.5),          # Center and set size for plot title
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 17), # Set size for x-axis text
+    axis.text.y = element_text(hjust = 1, size = 17),            # Set size for y-axis text
+    axis.title.x = element_text(size = 20, hjust = 0.5),         # Center and set size for x-axis title
+    axis.title.y = element_text(size = 20, hjust = 0.5),         # Center and set size for y-axis title
+  )
 
@@
 
-# Discussion
+The latest analysis of shelter occupancy data reveals a marked upward trend in average occupancy rates, rounded up at the first decimal place, over 21 days from January 1st to January 21st in 2024 (see @fig-average-occupancy-rate-by-date).
+While daily fluctuations were observed, the general direction was a steady increase in the need for shelter services.
+However, to provide a more comprehensive understanding, several critical aspects need to be incorporated into the discussion:
 
-## First discussion point {#sec-first-point}
+1.
+**Overall Rising Trend and Demographic Patterns**: The data shows an overall increase in occupancy rates.
+This trend aligns with the Demographic Patterns of Homelessness, suggesting a balanced need for both bed-based and room-based capacities in shelters.
+Notably, there's a pronounced requirement for bed-based capacity, particularly for family user groups.
+This insight is crucial for shelter resources to meet the specific needs of different homeless populations effectively.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+2.
+**Geographic Patterns of Homelessness and Occupancy Rate**: The occupancy rates across all listed cities are consistently high, indicating a widespread demand for shelter services (see @fig-average-occupancy-rate-by-city).
+This requires effective management of existing resources in these areas.
+Additionally, providing safe travel options to vacant shelters in other cities and access to education to go back to society for the homeless might be significant considerations.
+These services can play an important role in enhancing the quality of life for homeless individuals and families, and in some cases, may aid in reducing the incidence of homelessness.
 
-## Second discussion point
+3.
+**Possible Issues**: Despite the high occupancy rates, none of the shelters reached or exceeded 100% capacity.
+This observation is contrasted with the visible presence of many homeless individuals on the streets.
+This contradiction indicates potential issues with accessibility or awareness of shelter services.
+To address this gap, shelter operators and advocacy groups need to intensify their efforts.
+This could involve improving outreach programs, enhancing the accessibility of shelters, and removing barriers that prevent homeless individuals from seeking shelter.
+By doing so, more individuals can be brought under the protection of shelter programs, thereby reducing the number of homeless people without adequate shelter.
+
+## Limitations and Next Steps
+
+### Limitations
+1.
+**Data Timeframe and Recency**: One of the primary limitations of this study is the timeframe of the data analyzed.
+The 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] data contains the period from January 1st to January 21st, 2024.
+While the data has been updated daily since 2024, this analysis does not account for any updates or changes that may have occurred beyond January 21st.
+This restriction could potentially lead to minor inaccuracies or the omission of recent developments that could impact the findings and their relevance to the current situation.
+Additionally, the absence of data before 2024 limits the ability to observe significant trends and changes over a longer historical period.
+
+2.
+**Scope of Data Variables**: The analysis currently focuses on general occupancy rates without considering specific variables that could offer more detailed insights.
+For instance, the inclusion of variables such as the program model of shelters, the type of overnight services provided, or the program area could potentially yield more detailed and informative results.
+These factors can significantly influence shelter occupancy rates and patterns, and their exclusion from the analysis may limit the depth and applicability of the conclusions drawn.
+
+### Next Steps
+1.
+**Incorporation of Recent Data**: To address the limitation of the study's timeframe, future analyses should incorporate the most recent data containing a wider timeframe available.
+This approach will ensure that the findings are up-to-date and reflect the current state of shelter occupancy.
+Regular updates to the analysis will also help in monitoring trends over time, providing a dynamic view of the homelessness situation.
+
+2.
+**Expanding Data Parameters**: Future studies should aim to include additional variables such as the program model, service type, and program area of shelters.
+This expansion will allow for a more comprehensive analysis, potentially uncovering patterns and trends that are not evident from occupancy rates alone.
+For example, understanding how different program models affect occupancy rates can inform more targeted and effective policy decisions and resource allocation.
+
+In conclusion, while this study provides important insights into shelter occupancy trends, its scope and the recency of data are somewhat limited.
+Future research efforts should aim to overcome these limitations by incorporating more comprehensive and up-to-date data, thereby enriching the understanding of homelessness and informing more effective responses to this pressing social issue.
 

### 2024-01-25 outputs/paper/paper.qmd
update {#sec-discussion}

@@
 
-To dissect the demographic and geographic patterns within the shelter system in Toronto, spotlighting trends in service utilization and capacity, this paper is organized into the following sections: Data, and Discussion.
+To dissect the demographic and geographic patterns within the shelter system in Toronto, spotlighting trends in service utilization and capacity, this paper is organized into the following sections: Data (@sec-data), and Discussion (@sec-discussion).
 In the Data section, I discuss a detailed description of the data source and collection methods.
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-01-25 outputs/paper/paper.qmd
final revision

@@
 This study is designed for an in-depth analysis of Toronto's shelter occupancy and capacity, utilizing data [@ShelterData] from the Shelter Management Information System (SMIS) database.
-While there are extensive papers on homelessness, a clear gap exists in contemporary, data-driven analyses that integrate both demographic and geographic dimensions within the context of urban shelter systems.
-Previous studies have often been limited either in data scope or timeframe, failing to provide a comprehensive picture that reflects the current realities of urban homelessness.
+While there are several papers on homelessness, this paper more focus on data-driven analyses that integrate both demographic and geographic dimensions within the context of urban shelter systems.
+Other datasets have often been limited either in data scope or timeframe, failing to provide a comprehensive picture that reflects the current realities of urban homelessness.
 This paper addresses this gap better by offering an updated view of the shelter system in Toronto.
 
-This study analyzes shelter occupancy trends using a dataset [@ShelterData] from the Shelter Management Information System (SMIS) [@ShelterData], employing R [@r] statistical tools like `tidyverse` [@rTidyverse] and `ggplot2` [@rGgplot2].
-Findings reveal a heavy reliance on bed-based capacities mainly for non-family groups but not for families, alongside high occupancy rates across Toronto.
-These results are vital for guiding policymakers and social welfare organizations in enhancing shelter capacities and accessibility, and in reevaluating strategies to more effectively address urban homelessness.
+In detail, this study analyzes shelter occupancy trends using a dataset [@ShelterData] from the Shelter Management Information System (SMIS) [@ShelterData], employing R [@r] statistical tools like `tidyverse` [@rTidyverse] and `ggplot2` [@rGgplot2].
+Findings reveal and a heavy reliance on bed-based capacities mainly for non-family groups but not for families, alongside high occupancy rates across Toronto.
+These results are playing key roles for guiding policymakers and social welfare organizations in enhancing shelter capacities and accessibility, and in reevaluating strategies to more effectively address urban homelessness.
 
@@
 This research is based on a dataset sourced from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto].
-The specific dataset employed is named  [@ShelterData].
+The specific dataset employed is named 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData].
 This database acts as a detailed collection of daily entries related to active overnight shelter and associated services.
@@
 ## Data Collection
-In the process of selecting an appropriate data source for an in-depth analysis of current demographic and geographic trends within shelter services, several datasets were considered, including the 'Toronto Shelter System Flow data' [@ShelterSystemFlow] and the 'Daily Shelter Occupancy' [@DailyShelterOccupancy] datasets.
+In the process of selecting an appropriate data source for an in-depth analysis of current demographic and geographic trends within shelter services, several datasets were considered.
+Two other datasets considered are the 'Toronto Shelter System Flow data' [@ShelterSystemFlow] and the 'Daily Shelter Occupancy' [@DailyShelterOccupancy] datasets.
 However, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset was ultimately chosen for the following reasons:
@@
 2.
-The 'Daily Shelter Occupancy' [@DailyShelterOccupancy] dataset, although previously a pertinent source of information, has become less relevant due to its discontinuation.
+The 'Daily Shelter Occupancy' [@DailyShelterOccupancy] dataset, although previously a proper source of information, has become less relevant due to its discontinuation.
 The most recent update for this dataset was recorded on March 10, 2022, therefore it is outdated for current analysis purposes.
@@
 
-The methodology for gathering data is structured to ensure a detailed and current perspective of shelter and overnight service initiatives.
+The methodology for gathering data is structured to ensure a detailed and current perspective of shelter and overnight service.
 The data was compiled and examined using the R statistical programming software [@r], supplemented by various tools such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `knitr` [@rknitr], and `here` [@rHere] (see @tbl-cleaned-shelter-statistics).
@@
 
-A striking observation is the complete lack of bed-based capacity options for family group.
+A striking observation is the complete lack of bed-based capacity options for family groups.
 This suggests that the current shelter system does not offer bed-specific accommodations for families, possibly due to the unique needs and privacy concerns associated with family units.
@@
 While daily fluctuations were observed, the general direction was a steady increase in the need for shelter services.
-However, to provide a more comprehensive understanding, several critical aspects need to be incorporated into the discussion:
+However, to provide a more comprehensive understanding, several critical aspects need to be considered in the discussion:
 
@@
 **Overall Rising Trend and Demographic Patterns**: The data shows an overall increase in occupancy rates.
-This trend aligns with the Demographic Patterns of Homelessness, suggesting a balanced need for both bed-based and room-based capacities in shelters.
-Notably, there's a pronounced requirement for bed-based capacity, particularly for family user groups.
+This trend aligns with the demographic patterns of homelessness, suggesting a balanced need for both bed-based and room-based capacities in shelters.
+Notably, there's an obvious requirement for bed-based capacity, particularly for family user groups.
 This insight is crucial for shelter resources to meet the specific needs of different homeless populations effectively.
@@
 This requires effective management of existing resources in these areas.
-Additionally, providing safe travel options to vacant shelters in other cities and access to education to go back to society for the homeless might be significant considerations.
+Additionally, providing safe travel options to the vacant shelters in other cities and access to education to go back to society for the homeless might be significant considerations.
 These services can play an important role in enhancing the quality of life for homeless individuals and families, and in some cases, may aid in reducing the incidence of homelessness.
@@
 The 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] data contains the period from January 1st to January 21st, 2024.
-While the data has been updated daily since 2024, this analysis does not account for any updates or changes that may have occurred beyond January 21st.
+While the data has been updated daily since 2024, this analysis does not account for any updates or changes that may have occurred beyond January 21st, 2024.
 This restriction could potentially lead to minor inaccuracies or the omission of recent developments that could impact the findings and their relevance to the current situation.
@@
 For instance, the inclusion of variables such as the program model of shelters, the type of overnight services provided, or the program area could potentially yield more detailed and informative results.
-These factors can significantly influence shelter occupancy rates and patterns, and their exclusion from the analysis may limit the depth and applicability of the conclusions drawn.
+These factors can significantly influence shelter occupancy patterns, and their exclusion from the analysis may limit the depth and applicability of the conclusions drawn.
 
@@
 1.
-**Incorporation of Recent Data**: To address the limitation of the study's timeframe, future analyses should incorporate the most recent data containing a wider timeframe available.
+**Incorporation of Recent Data**: To address the limitation of the study's timeframe, future analyses should follow the most recent data update containing a wider timeframe available.
 This approach will ensure that the findings are up-to-date and reflect the current state of shelter occupancy.

### 2024-01-25 outputs/paper/paper.qmd
updated R project name

@@
 While there are several papers on homelessness, this paper more focus on data-driven analyses that integrate both demographic and geographic dimensions within the context of urban shelter systems.
-Other datasets have often been limited either in data scope or timeframe, failing to provide a comprehensive picture that reflects the current realities of urban homelessness.
+Other datasets have often been limited either in data scope or timeframe, failing to provide a picture that reflects the current realities of urban homelessness.
 This paper addresses this gap better by offering an updated view of the shelter system in Toronto.
@@
 The specific dataset employed is named 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData].
-This database acts as a detailed collection of daily entries related to active overnight shelter and associated services.
-It is the latest and most comprehensive dataset available for analyzing homelessness support services.
+This database acts as a detailed collection of daily entries related to active overnight shelters and associated services.
 
@@
 1.
-While the 'Toronto Shelter System Flow' [@ShelterSystemFlow] data offers valuable insights into the flow of demography through the shelter system, it lacks comprehensive geographic information.
+While the 'Toronto Shelter System Flow' [@ShelterSystemFlow] data offers valuable insights into the flow of demography through the shelter system, it lacks geographic information.
 This limitation significantly hinders the potential for a detailed analysis of geographic trends and patterns within the shelter system, an aspect critical to my research objectives.
@@
 The most recent update for this dataset was recorded on March 10, 2022, therefore it is outdated for current analysis purposes.
-In contrast, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset provides the most recent and comprehensive data available, making it a more suitable choice for analyzing up-to-date trends and developments in shelter usage and capacity.
+In contrast, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset provides the most recent and thorough data available, making it a more suitable choice for analyzing up-to-date trends and developments in shelter usage and capacity.
 
@@
 While daily fluctuations were observed, the general direction was a steady increase in the need for shelter services.
-However, to provide a more comprehensive understanding, several critical aspects need to be considered in the discussion:
+However, to provide a more thorough understanding, several critical aspects need to be considered in the discussion:
 
@@
 **Expanding Data Parameters**: Future studies should aim to include additional variables such as the program model, service type, and program area of shelters.
-This expansion will allow for a more comprehensive analysis, potentially uncovering patterns and trends that are not evident from occupancy rates alone.
+This expansion will allow for a more thorough analysis, potentially uncovering patterns and trends that are not evident from occupancy rates alone.
 For example, understanding how different program models affect occupancy rates can inform more targeted and effective policy decisions and resource allocation.
@@
 In conclusion, while this study provides important insights into shelter occupancy trends, its scope and the recency of data are somewhat limited.
-Future research efforts should aim to overcome these limitations by incorporating more comprehensive and up-to-date data, thereby enriching the understanding of homelessness and informing more effective responses to this pressing social issue.
+Future research efforts should aim to overcome these limitations by incorporating more thorough and up-to-date data, thereby enriching the understanding of homelessness and informing more effective responses to this pressing social issue.
 
